---
title: "LLE: Goose Bay Case Study"
author: "Jason Orender"
date: "2022-11-06"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(readr, quietly = TRUE)
```
## Introduction

This document was created by running the "Goose_Bay_markdown.Rmd" file in
RStudio. It utilizes the script files "functions.R" and "rectify.R" within the
"Code" directory in this repository, which are used to define the functions that
are called below.  This document reflects real world data, and walks through
the process from reading in the raw data to displaying the plots for
interpretation.

## About the Data

One thoroughly analyzed data set that is often used as an exemplar is the
ionosphere data collected by a system in Goose Bay, Labrador, downloaded from
the UCI data repository (<a href="http://archive.ics.uci.edu/ml/datasets/Ionosphere">UCI Data</a>). The original study was published in 1989 [<a href="https://www.jhuapl.edu/Content/techdigest/pdf/V10-N03/10-03-Sigillito_Class.pdf">33</a>] and has been subsequently used in at least 55 additional papers as a test set
for machine learning algorithms.  While this is not specifically longitudinal
data, it shows how the technique can be adapted to other data with similar
analysis impediments, and highlights an area that may be of particular interest:
disambiguating sensor returns to determine which sensors are producing relevant
data.

From the UCI page [<a href="http://archive.ics.uci.edu/ml/index.php">32</a>]:
"This radar data was collected by a system in Goose Bay, Labrador. This system
consists of a phased array of 16 high-frequency antennas with a total
transmitted power on the order of 6.4 kilowatts. See the <a href="https://www.jhuapl.edu/Content/techdigest/pdf/V10-N03/10-03-Sigillito_Class.pdf">paper</a> for more details. The targets were free electrons in the
ionosphere. "Good" radar returns are those showing evidence of some type of
structure in the ionosphere. "Bad" returns are those that do not; their signals
pass through the ionosphere.

Received signals were processed using an autocorrelation function whose
arguments are the time of a pulse and the pulse number. There were 17 pulse
numbers for the Goose Bay system. Instances in this databse are described by 2
attributes per pulse number, corresponding to the complex values returned by the
function resulting from the complex electromagnetic signal."

```{r load_functions, echo=FALSE}
# Loading the functions that will be used to process the data - these are
# defined in a separate script file to allow re-use between markdown documents
# as well as create a set of generalized code that might be useful to others.
source("../../Code/functions.R")
source("../../Code/rectify.R")
```

```{r load_data, echo=FALSE}
message("Reading in the data file...")
ionosphere_data <- data.frame(read_csv("ionosphere.data.csv", col_names = FALSE,
                                       show_col_types = FALSE))
# the data file did not have column names, so adding them
cnames <- c(unlist(lapply(1:17, FUN = function(i)
    { c(paste0("Pulse_", i, "_r"), paste0("Pulse_", i, "_i")) })), "class")
colnames(ionosphere_data) <- cnames
# transforming the response column into a logical vector
ionosphere_data$class <- ionosphere_data$class == "g"
ionosphere_data$X     <- 1:dim(ionosphere_data)[1]

# setting the names of the response column and the excluded columns
resp    <- "class"
exclude <- c("X")

# naming the distinct groups within the data
ion_groups  <- list()
ion_groups[['Pulse_r']] <- paste0("Pulse_", 1:17, "_r")
ion_groups[['Pulse_i']] <- paste0("Pulse_", 1:17, "_i")

# creating a character vector of all features
feats               <- unlist(ion_groups)

message("  Done.")

# splitting the training and testing data
message("Splitting the data...")
rseed       <- 2345 # using a random seed for repeatability
train_split <- 0.67
pos_balance <- 1.0
neg_balance <- 1.0

# the entire data set
ion       <- ionosphere_data
# splitting the positive and negative examples so that we can control how many
# of each go into the training and testing sets, rather than leave that to
# chance
ion_pos   <- ionosphere_data[ionosphere_data$class,]
ion_neg   <- ionosphere_data[!ionosphere_data$class,]

# setting the seed
set.seed(rseed)
# taking a random sample of the positive and negative examples for the training
# set
ion_train <- rbind(ion_pos[sample(dim(ion_pos)[1], train_split*pos_balance*dim(ion_pos)[1]),],
                   ion_neg[sample(dim(ion_neg)[1], train_split*neg_balance*dim(ion_neg)[1]),])
# taking everything that's left and putting it in the testing set
ion_test  <- rbind(ion_pos[!(ion_pos$X %in% ion_train$X),],
                   ion_neg[!(ion_neg$X %in% ion_train$X),])
message("  Done.")
```

## Studying the Goose Bay Data

```{r fit_models, echo=FALSE}
# This code block uses a specialized routine called "modelfit" which is a
# generalized routine that fits data to a specified model type.  It was made
# to simplify the interface and make the code more succinct.
message("Fitting Models...")
ion_sq_fit <- modelfit(data = ion_train, exclude = exclude, response = resp,
                       fit_type = "SQ", groups = ion_groups,
                       params = list(sdfilter=NULL))
# Fitting the LASSO on the untransformed data
ion_ls_fit <- modelfit(data = ion_train, fit_type = "LS", exclude = exclude,
                       response = resp, groups = ion_groups)
message("  Done.")
```
The LASSO with transformed data (shown immediately below) again exceeded the
performance the LASSO with the original un-transformed data (shown below the
figures for the transformed data) and nearly matched the neural network that was
specifically tuned and used to analyze the data in the original paper.  Note
also the similar training and testing results for the transformed data, while
there is a large disparity between the training and testing results for the un-
transformed data.  This is an example of how the transformed data resists
overfitting.

```{r transformed_data_plots, echo=FALSE}
hpos <- 0.6
plot(ion_sq_fit, title = "Goose Bay Training Set (transformed data)", h=.5,
     cx = hpos, ylab = "Predicted Probability")
plot(ion_sq_fit, data = ion_test, h=.5, cx = hpos, cy = .4,
     title = "Goose Bay Testing Set (transformed data)",
     ylab = "Predicted Probability")
```

```{r untransformed_data_plots, echo=FALSE}
plot(ion_ls_fit, title = "Goose Bay Training Set (un-transformed data)", h=.5,
     cx = hpos, ylab = "Predicted Probability")
plot(ion_ls_fit, data = ion_test, h=.5, cx = hpos,
     title = "Goose Bay Testing Set (un-transformed data)",
     ylab = "Predicted Probability")
```

Additionally, a sparse coefficient vector was returned (shown below), which
clearly indicates which signals are important, along with the critical ranges of
the important signals, which are arguably nearly as interesting as the model
results alone.

```{r transformed_data_barplot, echo=FALSE}
BARHT   <- 5
beta1   <- ion_sq_fit$model$beta[,]
beta1   <- beta1[sort(names(beta1))]
markers <- rep(0,length(beta1))

par(new=FALSE, mar = c(6,3,3,2))
barplot(beta1, border = "blue", col = "blue", las = 2, xaxt = "n", ylim = c(0,5))
par(new=TRUE)
bplot <- barplot(markers, border = "blue", col = "blue",
                 yaxt = "n", las=2, ylim = c(0,5),
                 main = "Coefficients from Transformed Data")
cnames <- c("Pulse_2_r", "Pulse_3_r", "Pulse_2_i", "Pulse_3_i", "Pulse_4_i",
            "Pulse_7_i", "Pulse_9_i")
bplot_at <- NULL
for (i in 1:length(cnames))
  bplot_at <- c(bplot_at,grep(paste0(cnames[i],"$"),names(beta1)))
bplot_labels <- rep(NA, length(beta1))
bplot_labels[bplot_at] <- cnames
axis(1, at = bplot, labels =bplot_labels, las=2, cex.axis=0.8, tick=FALSE)

```

The coefficient for the LASSO solution on the un-transformed data (see below),
by contrast, shows no clear feature attribution.  The largest coefficient, in
fact, seems to be one which as no clear connection to the solution at all.

```{r untransformed_data_barplot, echo=FALSE}
beta1   <- ion_ls_fit$model$beta[,]
markers <- rep(0,length(beta1))

par(new=FALSE, mar = c(6,3,3,2))
barplot(beta1, border = "blue", col = "blue", las = 2, main = "Coefficients from Un-Transformed Data")
```

